---
title: "🌿 NDVI over Oʻahu with Sentinel-2 (R Tutorial)"
subtitle: "From Copernicus Browser to NDVI map in minutes"
author: "Your Name"
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "On this page"
    toc-depth: 3
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r}
library(here)
library(terra)
library(sf)
library(leaflet)
library(mapview)
library(dplyr)
library(plotly)
```

## NDVI (Normalized Difference Vegetation Index)

NDVI is a simple, per-pixel indicator of “greenness” (photosynthetic activity).

### What it measures

Plants reflect more **near-infrared (NIR)** light and absorb **red** light. NDVI combines those two bands to highlight living vegetation.

### Formula

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{=html}
<!-- Plain-text fallback (if math is disabled):
NDVI = (NIR - Red) / (NIR + Red)
-->
```

### Typical value ranges

-   **0.6–0.9**: dense, healthy vegetation\
-   **0.2–0.5**: sparse/brown vegetation, shrubs/grass\
-   **0–0.2**: bare soil / urban\
-   **\< 0**: water, clouds, shadows, snow

### What data do we need to calculate NDVI

-   We can use the [Copernicus Open Access Hub](https://browser.dataspace.copernicus.eu/) to grab data form the satellite Sentinel 2

-   We specifically need Band 8 data for near-infared and Band 4 data for red

```{r}
# Paths to your .jp2 files
red_path2019 <- here("data/kealakekua/2019/T04QHG_20190116T210921_B04_10m.jp2")
nir_path2019 <- here("data/kealakekua/2019/T04QHG_20190116T210921_B08_10m.jp2")

red_path2020 <- here("data/kealakekua/2020/T04QHG_20200225T210919_B04_10m.jp2")
nir_path2020 <- here("data/kealakekua/2020/T04QHG_20200225T210919_B08_10m.jp2")

red_path2021 <- here("data/kealakekua/2021/T04QHG_20211116T210919_B04_10m.jp2")
nir_path2021 <- here("data/kealakekua/2021/T04QHG_20211116T210919_B08_10m.jp2")

red_path2022 <- here("data/kealakekua/2022/T04QHG_20220301T210931_B04_10m.jp2")
nir_path2022 <- here("data/kealakekua/2022/T04QHG_20220301T210931_B08_10m.jp2")

red_path2023 <- here("data/kealakekua/2023/T04QHG_20230306T210921_B04_10m.jp2")
nir_path2023 <- here("data/kealakekua/2023/T04QHG_20230306T210921_B08_10m.jp2")

red_path2024 <- here("data/kealakekua/2024/T04QHG_20240214T210919_B04_10m.jp2")
nir_path2024 <- here("data/kealakekua/2024/T04QHG_20240214T210919_B08_10m.jp2")
```

### These .jp2 files contain raster data

-   Raster data is a way to represent geography as a **grid of cells (pixels)**. Each cell has a single value (e.g., elevation, temperature, NDVI) and the whole grid is georeferenced so every cell aligns to a real-world location

-   The `terra` package helps us read raster data into R

```{r}
#Read bands directly
red2019 <- rast(red_path2019)
nir2019 <- rast(nir_path2019)

red2020 <- rast(red_path2020)
nir2020 <- rast(nir_path2020)

red2021 <- rast(red_path2021)
nir2021 <- rast(nir_path2021)

red2022 <- rast(red_path2022)
nir2022 <- rast(nir_path2022)

red2023 <- rast(red_path2023)
nir2023 <- rast(nir_path2023)

red2024 <- rast(red_path2024)
nir2024 <- rast(nir_path2024)
```

### Calculating NDVI

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{r}
ndvi2019 <- (nir2019 - red2019) / (nir2019 + red2019)

ndvi2020 <- (nir2020 - red2020) / (nir2020 + red2020)

ndvi2021 <- (nir2021 - red2021) / (nir2021 + red2021)

ndvi2022 <- (nir2022 - red2022) / (nir2022 + red2022)

ndvi2023 <- (nir2023 - red2023) / (nir2023 + red2023)

ndvi2024 <- (nir2024 - red2024) / (nir2024 + red2024)
```

### Plotting NDVI

```{r}
plot(ndvi2019, main = "NDVI 2019")

plot(ndvi2020, main = "NDVI 2020")

plot(ndvi2021, main = "NDVI 2021")

plot(ndvi2022, main = "NDVI 2022")

plot(ndvi2023, main = "NDVI 2023")

plot(ndvi2024, main = "NDVI 2024")
```

```{r}
ndvi19 <- aggregate(ndvi2019, fact = 4, fun = mean)
ndvi20 <- aggregate(ndvi2020, fact = 4, fun = mean)
ndvi21 <- aggregate(ndvi2021, fact = 4, fun = mean)
ndvi22 <- aggregate(ndvi2022, fact = 4, fun = mean)
ndvi23 <- aggregate(ndvi2023, fact = 4, fun = mean)
ndvi24 <- aggregate(ndvi2024, fact = 4, fun = mean)

# Create custom breaks and colors
breaks <- c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1)
colors <- c("#4A90E2",      # Ocean/Snow/Cloud (blue)
           "#CD853F",       # Bare soil/urban (tan)
           "#DEB887",       # Very sparse (light tan)
           "#F0E68C",       # Sparse (light yellow)
           "#ADFF2F",       # Moderate (yellow-green)
           "#32CD32",       # Good vegetation (lime green)
           "#228B22",       # Dense vegetation (forest green)
           "#006400")       # Very dense (dark green)

pal <- colorBin(palette = colors,
                bins = breaks,
                na.color = "transparent",
                right = FALSE)

mapview(ndvi19, zcol = "layer", at = breaks, col.regions = colors, layer.name = "NDVI",
        alpha = 0.8)
mapview(ndvi20, zcol = "layer", at = breaks, col.regions = colors, layer.name = "NDVI",
        alpha = 0.8)
mapview(ndvi21, zcol = "layer", at = breaks, col.regions = colors, layer.name = "NDVI",
        alpha = 0.8)
mapview(ndvi22, zcol = "layer", at = breaks, col.regions = colors, layer.name = "NDVI",
        alpha = 0.8)
mapview(ndvi23, zcol = "layer", at = breaks, col.regions = colors, layer.name = "NDVI",
        alpha = 0.8)
mapview(ndvi24, zcol = "layer", at = breaks, col.regions = colors, layer.name = "NDVI",
        alpha = 0.8)
```

```{r}
# Use the reduced-resolution raster
ndvi19_values <- values(ndvi19)
ndvi19_values <- ndvi19_values[!is.na(ndvi19_values)]  # Remove NA

# Define NDVI categories
ndvi19_categories <- cut(ndvi19_values,
                         breaks = c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1),
                         labels = c("Water/Snow/Cloud (-1–0)",
                                    "Bare Soil / Urban (0–0.1)",
                                    "Very Sparse Vegetation (0.1–0.2)",
                                    "Sparse Vegetation (0.2–0.3)",
                                    "Moderate Vegetation (0.3–0.4)",
                                    "Good Vegetation (0.4–0.5)",
                                    "Dense Vegetation (0.5–0.6)",
                                    "Very Dense Vegetation (0.6–1)"),
                         right = FALSE)


# Create data frame with counts
ndvi19_df <- as.data.frame(table(ndvi19_categories))
colnames(ndvi19_df) <- c("Category", "Count")

# Calculate percentages
ndvi19_df <- ndvi19_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi19_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages - 2019",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",      # Ocean/Snow/Cloud (blue)
           "#CD853F",       # Bare soil/urban (tan)
           "#DEB887",       # Very sparse (light tan)
           "#F0E68C",       # Sparse (light yellow)
           "#ADFF2F",       # Moderate (yellow-green)
           "#32CD32",       # Good vegetation (lime green)
           "#228B22",       # Dense vegetation (forest green)
           "#006400")       # Very dense (dark green)
  ) +
  theme_minimal() +
  coord_flip()
```

```{r}
# Use the reduced-resolution raster
ndvi20_values <- values(ndvi20)
ndvi20_values <- ndvi20_values[!is.na(ndvi20_values)]  # Remove NA

# Define NDVI categories
ndvi20_categories <- cut(ndvi20_values,
                         breaks = c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1),
                         labels = c("Water/Snow/Cloud (-1–0)",
                                    "Bare Soil / Urban (0–0.1)",
                                    "Very Sparse Vegetation (0.1–0.2)",
                                    "Sparse Vegetation (0.2–0.3)",
                                    "Moderate Vegetation (0.3–0.4)",
                                    "Good Vegetation (0.4–0.5)",
                                    "Dense Vegetation (0.5–0.6)",
                                    "Very Dense Vegetation (0.6–1)"),
                         right = FALSE)

# Create data frame with counts
ndvi20_df <- as.data.frame(table(ndvi20_categories))
colnames(ndvi20_df) <- c("Category", "Count")

# Calculate percentages
ndvi20_df <- ndvi20_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi20_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages - 2020",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",  # Ocean/Snow/Cloud (blue)
    "#CD853F",  # Bare soil/urban (tan)
    "#DEB887",  # Very sparse (light tan)
    "#F0E68C",  # Sparse (light yellow)
    "#ADFF2F",  # Moderate (yellow-green)
    "#32CD32",  # Good vegetation (lime green)
    "#228B22",  # Dense vegetation (forest green)
    "#006400"   # Very dense (dark green)
  )) +
  theme_minimal() +
  coord_flip()

```

```{r}
# Use the reduced-resolution raster
ndvi21_values <- values(ndvi21)
ndvi21_values <- ndvi21_values[!is.na(ndvi21_values)]  # Remove NA

# Define NDVI categories
ndvi21_categories <- cut(ndvi21_values,
                         breaks = c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1),
                         labels = c("Water/Snow/Cloud (-1–0)",
                                    "Bare Soil / Urban (0–0.1)",
                                    "Very Sparse Vegetation (0.1–0.2)",
                                    "Sparse Vegetation (0.2–0.3)",
                                    "Moderate Vegetation (0.3–0.4)",
                                    "Good Vegetation (0.4–0.5)",
                                    "Dense Vegetation (0.5–0.6)",
                                    "Very Dense Vegetation (0.6–1)"),
                         right = FALSE)

# Create data frame with counts
ndvi21_df <- as.data.frame(table(ndvi21_categories))
colnames(ndvi21_df) <- c("Category", "Count")

# Calculate percentages
ndvi21_df <- ndvi21_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi21_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages - 2021",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",  # Ocean/Snow/Cloud (blue)
    "#CD853F",  # Bare soil/urban (tan)
    "#DEB887",  # Very sparse (light tan)
    "#F0E68C",  # Sparse (light yellow)
    "#ADFF2F",  # Moderate (yellow-green)
    "#32CD32",  # Good vegetation (lime green)
    "#228B22",  # Dense vegetation (forest green)
    "#006400"   # Very dense (dark green)
  )) +
  theme_minimal() +
  coord_flip()

```

```{r}
# Use the reduced-resolution raster
ndvi22_values <- values(ndvi22)
ndvi22_values <- ndvi22_values[!is.na(ndvi22_values)]  # Remove NA

# Define NDVI categories
ndvi22_categories <- cut(ndvi22_values,
                         breaks = c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1),
                         labels = c("Water/Snow/Cloud (-1–0)",
                                    "Bare Soil / Urban (0–0.1)",
                                    "Very Sparse Vegetation (0.1–0.2)",
                                    "Sparse Vegetation (0.2–0.3)",
                                    "Moderate Vegetation (0.3–0.4)",
                                    "Good Vegetation (0.4–0.5)",
                                    "Dense Vegetation (0.5–0.6)",
                                    "Very Dense Vegetation (0.6–1)"),
                         right = FALSE)

# Create data frame with counts
ndvi22_df <- as.data.frame(table(ndvi22_categories))
colnames(ndvi22_df) <- c("Category", "Count")

# Calculate percentages
ndvi22_df <- ndvi22_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi22_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages - 2022",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",  # Ocean/Snow/Cloud (blue)
    "#CD853F",  # Bare soil/urban (tan)
    "#DEB887",  # Very sparse (light tan)
    "#F0E68C",  # Sparse (light yellow)
    "#ADFF2F",  # Moderate (yellow-green)
    "#32CD32",  # Good vegetation (lime green)
    "#228B22",  # Dense vegetation (forest green)
    "#006400"   # Very dense (dark green)
  )) +
  theme_minimal() +
  coord_flip()

```

```{r}
# Use the reduced-resolution raster
ndvi23_values <- values(ndvi23)
ndvi23_values <- ndvi23_values[!is.na(ndvi23_values)]  # Remove NA

# Define NDVI categories
ndvi23_categories <- cut(ndvi23_values,
                         breaks = c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1),
                         labels = c("Water/Snow/Cloud (-1–0)",
                                    "Bare Soil / Urban (0–0.1)",
                                    "Very Sparse Vegetation (0.1–0.2)",
                                    "Sparse Vegetation (0.2–0.3)",
                                    "Moderate Vegetation (0.3–0.4)",
                                    "Good Vegetation (0.4–0.5)",
                                    "Dense Vegetation (0.5–0.6)",
                                    "Very Dense Vegetation (0.6–1)"),
                         right = FALSE)

# Create data frame with counts
ndvi23_df <- as.data.frame(table(ndvi23_categories))
colnames(ndvi23_df) <- c("Category", "Count")

# Calculate percentages
ndvi23_df <- ndvi23_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi23_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages - 2023",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",  # Ocean/Snow/Cloud (blue)
    "#CD853F",  # Bare soil/urban (tan)
    "#DEB887",  # Very sparse (light tan)
    "#F0E68C",  # Sparse (light yellow)
    "#ADFF2F",  # Moderate (yellow-green)
    "#32CD32",  # Good vegetation (lime green)
    "#228B22",  # Dense vegetation (forest green)
    "#006400"   # Very dense (dark green)
  )) +
  theme_minimal() +
  coord_flip()

```

```{r}
# Use the reduced-resolution raster
ndvi24_values <- values(ndvi24)
ndvi24_values <- ndvi24_values[!is.na(ndvi24_values)]  # Remove NA

# Define NDVI categories
ndvi24_categories <- cut(ndvi24_values,
                         breaks = c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1),
                         labels = c("Water/Snow/Cloud (-1–0)",
                                    "Bare Soil / Urban (0–0.1)",
                                    "Very Sparse Vegetation (0.1–0.2)",
                                    "Sparse Vegetation (0.2–0.3)",
                                    "Moderate Vegetation (0.3–0.4)",
                                    "Good Vegetation (0.4–0.5)",
                                    "Dense Vegetation (0.5–0.6)",
                                    "Very Dense Vegetation (0.6–1)"),
                         right = FALSE)

# Create data frame with counts
ndvi24_df <- as.data.frame(table(ndvi24_categories))
colnames(ndvi24_df) <- c("Category", "Count")

# Calculate percentages
ndvi24_df <- ndvi24_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi24_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages - 2024",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",  # Ocean/Snow/Cloud (blue)
    "#CD853F",  # Bare soil/urban (tan)
    "#DEB887",  # Very sparse (light tan)
    "#F0E68C",  # Sparse (light yellow)
    "#ADFF2F",  # Moderate (yellow-green)
    "#32CD32",  # Good vegetation (lime green)
    "#228B22",  # Dense vegetation (forest green)
    "#006400"   # Very dense (dark green)
  )) +
  theme_minimal() +
  coord_flip()

```
