---
title: "🌿 NDVI over Oʻahu with Sentinel-2 (R Tutorial)"
subtitle: "From Copernicus Browser to NDVI map in minutes"
author: "Your Name"
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "On this page"
    toc-depth: 3
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r}
library(here)
library(terra)
library(sf)
library(leaflet)
library(mapview)
library(dplyr)
library(plotly)
```

## NDVI (Normalized Difference Vegetation Index)

NDVI is a simple, per-pixel indicator of “greenness” (photosynthetic activity).

### What it measures

Plants reflect more **near-infrared (NIR)** light and absorb **red** light. NDVI combines those two bands to highlight living vegetation.

### Formula

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{=html}
<!-- Plain-text fallback (if math is disabled):
NDVI = (NIR - Red) / (NIR + Red)
-->
```

### Typical value ranges

-   **0.6–0.9**: dense, healthy vegetation\
-   **0.2–0.5**: sparse/brown vegetation, shrubs/grass\
-   **0–0.2**: bare soil / urban\
-   **\< 0**: water, clouds, shadows, snow

### What data do we need to calculate NDVI

-   We can use the [Copernicus Open Access Hub](https://browser.dataspace.copernicus.eu/) to grab data form the satellite Sentinel 2

-   We specifically need Band 8 data for near-infared and Band 4 data for red

```{r}
# Paths to your .jp2 files
red_path <- here("data/kealakekua/T05QKB_20190908T210929_B04_10m.jp2")
nir_path <- here("data/kealakekua/T05QKB_20190908T210929_B08_10m.jp2")
```

### These .jp2 files contain raster data

-   Raster data is a way to represent geography as a **grid of cells (pixels)**. Each cell has a single value (e.g., elevation, temperature, NDVI) and the whole grid is georeferenced so every cell aligns to a real-world location

-   The `terra` package helps us read raster data into R

```{r}
# Read bands directly
red <- rast(red_path)
nir <- rast(nir_path)
```

### Calculating NDVI

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{r}
ndvi <- (nir - red) / (nir + red)
```

### Plotting NDVI

```{r}
plot(ndvi, main = "NDVI")
```

```{r}
# Reduce resolution significantly (factor of 8-10 should work)
ndvi_adv <- aggregate(ndvi, fact = 4, fun = mean)

# Create custom breaks and colors
breaks <- c(-1, 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 1)
colors <- c("#4A90E2",      # Ocean (blue)
           "#CD853F",       # Bare soil/urban (tan)
           "#DEB887",       # Very sparse (light tan)
           "#F0E68C",       # Sparse (light yellow)
           "#ADFF2F",       # Moderate (yellow-green)
           "#32CD32",       # Good vegetation (lime green)
           "#228B22",       # Dense vegetation (forest green)
           "#006400")       # Very dense (dark green)

pal <- colorBin(palette = colors,
                bins = breaks,
                na.color = "transparent",
                right = FALSE)

mapview(ndvi_adv,
        zcol = "layer",
        at = breaks,
        col.regions = colors,
        layer.name = "NDVI",
        alpha = 0.8)
```

```{r}
# Use the reduced-resolution raster
ndvi_values <- values(ndvi_adv)
ndvi_values <- ndvi_values[!is.na(ndvi_values)]  # Remove NA

# Define simplified NDVI categories
ndvi_categories <- cut(ndvi_values,
                       breaks = c(-Inf, 0, 0.2, 0.5, 0.9),
                       labels = c("Water/Clouds/Snow (<0)",
                                  "Bare Soil / Urban (0–0.2)",
                                  "Sparse/Brown Vegetation (0.2–0.5)",
                                  "Dense Healthy Vegetation (0.6–0.9)"),
                       right = FALSE)

# Create data frame with counts
ndvi_df <- as.data.frame(table(ndvi_categories))
colnames(ndvi_df) <- c("Category", "Count")

# Calculate percentages
ndvi_df <- ndvi_df %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot
ggplot(ndvi_df, aes(x = Category, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "NDVI Category Percentages",
       x = "NDVI Category",
       y = "Percentage (%)") +
  scale_fill_manual(values = c(
    "#4A90E2",  # Water
    "#CD853F",  # Bare soil / urban
    "#F0E68C",  # Sparse vegetation
    "#228B22"   # Dense vegetation
  )) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  coord_flip()
```
